Decoding Sleep Disorders Through Self-Reported Patterns

### 2024-04-17 paper/paper.qmd
added simulation

---
title: "Decoding Sleep Disorders Through Self-Reported Patterns"
subtitle: "A Logistic Regression Approach to the NHANES 2017-March 2020 Sleep Data"
title-block-banner: true
abstract: "This study investigates the potential relationship between individuals' sleep patterns and the self-reported incidence of sleep troubles as confirmed by a medical professional. Using logistic regression analysis on data from the NHANES 2017-March 2020 Sleep Disorders dataset, we examine how various factors such as sleep duration on weekdays and weekends, frequency of snoring, and daytime sleepiness are associated with the likelihood of reporting sleep problems to a doctor. The findings aim to shed light on the predictive value of self-monitored sleep behaviors in identifying individuals who may require medical attention for sleep-related issues. This analysis contributes to the broader understanding of sleep health and its complex interactions with daily functioning."
thanks: "Code and data supporting this analysis are available at https://github.com/TEJMaster/Sleep-Disorder-Analysis"
author: "Terry Tu"
date: today
date-format: long
format: pdf
toc: true
toc-depth: 1
number-sections: true
bibliography: references.bib
link-citations: true
editor: 
  markdown: 
    wrap: sentence
---

@@

\begin{align}
y_i | p_i &\sim \mbox{Bernoulli}(p_i) \\
\mbox{logit}(p_i) &= \beta_0 + \beta_1 \times x_{i1} + \beta_2 \times x_{i2} + \beta_3 \times x_{i3} + \beta_4 \times x_{i4} + \beta_5 \times x_{i5} \\
\beta_0 &\sim \mbox{Normal}(0, 1) \\ % Assuming the log-odds of the base rate of reported sleep trouble is centered around 0
\beta_1, \beta_2, \beta_3, \beta_4, \beta_5 &\sim \mbox{Normal}(0, 1) \\ % Weekday sleep duration effect and Weekend sleep duration effect, Snoring frequency,  Breathing pause frequency, and Overly sleep frequency
\end{align}

The logistic regression model is defined using a Bayesian framework, implemented in R with the `rstanarm` package. This approach allows us to incorporate prior knowledge about the parameters and to estimate their posterior distributions based on the observed data.

@@

The priors for the intercept ($\beta_0$) and the slopes ($\beta_1$, $\beta_2$, $\beta_3$, $\beta_4$, $\beta_5$) are deliberately chosen to reflect initial neutrality or skepticism regarding the impact of each predictor on the likelihood of reporting sleep troubles. Specifically, $\beta_0$ is assigned a Normal(0, 1) prior, indicating our initial uncertainty about the base rate of sleep trouble in the population.

For $\beta_1$ and $\beta_2$, which correspond to the effects of weekday and weekend sleep duration, we posit Normal(0, 0.5) priors. These priors are centered around zero, reflecting the number of sleep hour can be either positively or negatively contribute to sleep troubles, with a modest expectation of effect size.

The coefficients $\beta_3$, $\beta_4$, $\beta_5$ relating to snoring frequency, breathing pause frequency, and overly sleep frequency are each given a Normal(0, 1) prior. These priors imply that, prior to analyzing the data, we do not have strong expectations about the direction or scale of these effects.

@@

**Weekday Sleep Duration ($\beta_1$)**: We initially hypothesize that there may be a complex relationship between sleep duration on weekdays and sleep troubles. Long sleep durations could indicate an attempt to compensate for poor quality sleep, potentially related to sleep disturbances. As such, the prior for this coefficient is centered around zero, allowing the data to indicate the direction of association.

**Weekend Sleep Duration ($\beta_2$)**: For sleep duration on weekends, we consider the possibility of a recuperative effect, where catching up on sleep may reduce the likelihood of reporting troubles. This hypothesis is tentative, hence a prior centered around zero with a conservative standard deviation, permitting data-driven insights into this relationship.

**Snoring Frequency ($\beta_3$)**: Frequent snoring is a recognized symptom of sleep disorders like obstructive sleep apnea, which can lead to disrupted sleep. We posit a positive association between snoring frequency and reported sleep troubles, reflected in the priors for this coefficient.

**Breathing Pause Frequency ($\beta_4$)**: Breathing pauses during sleep are indicative of sleep apnea, which is directly related to sleep troubles. We hypothesize a positive association for this coefficient, which is incorporated into the prior setting.

**Daytime Sleepiness ($\beta_5$)**: Excessive daytime sleepiness is often the result of inadequate or disrupted nighttime sleep. We hypothesize a positive relationship between daytime sleepiness and the reporting of sleep troubles, expecting that those experiencing higher levels of sleepiness will be more likely to report troubles.

@@

# Print the summary table
coefficient_table
```
@tbl-logistic-coefficients presents the key coefficients for understanding the influence of various sleep-related variables on the likelihood of reporting sleep troubles:

**Intercept ($\beta_0$)**:
The intercept of the model, $\beta_0$, is estimated at -1.396. This value represents the log odds of reporting sleep troubles when all the predictors are held at zero. It suggests that, in the absence of any sleep issues or without considering any sleep behaviors, the baseline log odds of reporting sleep troubles to a doctor are negative, indicating a lower probability of reporting sleep troubles.

**Weekday Sleep Duration ($\beta_1$)**:
For the weekday sleep duration, the coefficient $\beta_1$ is 0.051, suggesting a slight increase in the log odds of reporting sleep troubles for each additional hour of sleep during the weekdays. This could be indicative of a situation where those who sleep more on weekdays might be doing so due to sleep issues that have led them to consult with a doctor.

**Weekend Sleep Duration ($\beta_2$)**:
Conversely, the coefficient for weekend sleep duration, $\beta_2$, is -0.131. This implies that for each additional hour of sleep during the weekends, there is a decrease in the log odds of reporting sleep troubles. It could be interpreted as those who manage to sleep more on weekends are less likely to report sleep problems, possibly because catching up on sleep may alleviate some of their weekday sleep deficits.

**Snoring Frequency ($\beta_3$)**:
The coefficient for snoring frequency, $\beta_3$, has a value of 0.032. While this is a positive value, suggesting that an increase in snoring frequency is associated with an increase in the likelihood of reporting sleep troubles, the effect size is relatively small.

**Breathing Pause Frequency ($\beta_4$)**:
The coefficient for breathing pause frequency, $\beta_4$, is 0.412, indicating a more substantial positive association with reporting sleep troubles. This aligns with clinical understanding, as breathing pauses are often associated with sleep disorders like sleep apnea, which can be a significant concern prompting medical consultation.

**Overly Sleep Frequency ($\beta_5$)**:
Lastly, the coefficient for overly sleep frequency, $\beta_5$, is 0.425, suggesting a strong relationship with the reporting of sleep troubles. This result is intuitive, as feeling excessively sleepy can be a direct symptom of poor sleep quality or a sleep disorder, leading to discussions with a healthcare provider.

@@

\begin{align}
\text{logit}(p_i) &= -1.396 + 0.051 \times x_{i1} - 0.131 \times x_{i2} + 0.032 \times x_{i3} + 0.412 \times x_{i4} + 0.425 \times x_{i5} + \varepsilon_i
\end{align}

## Predict Sleep Disorder Through Self Reported Patterns

@@

## Analysis of Residuals {#sec-residuals}

In this section, we evaluate the residuals from our logistic regression model to assess the model's goodness of fit and identify any potential anomalies in our data.

```{r residual-plots, echo=FALSE}
predicted_probs <- predict(bayesian_logistic_model, type = "response")
residuals <- residuals(bayesian_logistic_model, type = "pearson")

# Plotting residuals
plot(predicted_probs, residuals, xlab = "Predicted Probabilities", ylab = "Standardized Residuals")
abline(h = 0, col = "red")
```
## Model Performance

@@

# Appendix {.unnumbered}

@@

>>>>

---
title: "Decoding Sleep Disorders Through Self-Reported Patterns"
subtitle: "A Logistic Regression Approach to the NHANES 2017-March 2020 Sleep Data"
title-block-banner: true
abstract: "This study investigates the potential relationship between individuals' sleep patterns and the self-reported incidence of sleep troubles as confirmed by a medical professional. Using logistic regression analysis on data from the NHANES 2017-March 2020 Sleep Disorders dataset, we examine how various factors such as sleep duration on weekdays and weekends, frequency of snoring, and daytime sleepiness are associated with the likelihood of reporting sleep problems to a doctor. The results revealed that certain behaviors such as excessive snoring and high levels of daytime sleepiness showed a positive correlation with sleep troubles, Moreover, the duration of sleep showed a contrasting pattern of association for sleep during weekdays and weekends. This analysis contributes to the broader understanding of sleep health and its complex interactions with daily functioning."
thanks: "Code and data supporting this analysis are available at https://github.com/TEJMaster/Sleep-Disorder-Analysis"
author: "Terry Tu"
date: today
date-format: long
format: pdf
toc: true
toc-depth: 1
number-sections: true
bibliography: references.bib
link-citations: true
editor: 
  markdown: 
    wrap: sentence
---

@@

```{=tex}
\begin{align}
y_i | p_i &\sim \mbox{Bernoulli}(p_i) \\
\mbox{logit}(p_i) &= \beta_0 + \beta_1 \times x_{i1} + \beta_2 \times x_{i2} + \beta_3 \times x_{i3} + \beta_4 \times x_{i4} + \beta_5 \times x_{i5} \\
\beta_0 &\sim \mbox{Normal}(0, 1) \\ % Assuming the log-odds of the base rate of reported sleep trouble is centered around 0
\beta_1, \beta_2, \beta_3, \beta_4, \beta_5 &\sim \mbox{Normal}(0, 1) \\ % Weekday sleep duration effect and Weekend sleep duration effect, Snoring frequency,  Breathing pause frequency, and Overly sleep frequency
\end{align}
```
The logistic regression model is defined using a Bayesian framework, implemented in R with the `rstanarm` package.
This approach allows us to incorporate prior knowledge about the parameters and to estimate their posterior distributions based on the observed data.

@@

The priors for the intercept ($\beta_0$) and the slopes ($\beta_1$, $\beta_2$, $\beta_3$, $\beta_4$, $\beta_5$) are deliberately chosen to reflect initial neutrality or skepticism regarding the impact of each predictor on the likelihood of reporting sleep troubles.

Specifically, the intercept $\beta_0$ and the coefficients for weekday and weekend sleep duration, $\beta_1$ and $\beta_2$, are assigned Normal(0, 1) priors to maintain an unbiased stance before observing the data. This reflects an initial uncertainty about the base rate of sleep trouble in the population and acknowledges that the number of sleep hours could either positively or negatively influence sleep troubles, with only a modest expectation of effect size.

For the predictors related to snoring frequency, breathing pause frequency, and overly sleep frequency—$\beta_3$, $\beta_4$, $\beta_5$—the same Normal(0, 1) priors are utilized, which suggests that there are no preconceived notions regarding the magnitude or direction of their impact on reporting sleep troubles. This choice of priors allows the data to guide the inference about these predictors without the influence of strong initial assumptions.

@@

**Weekday Sleep Duration (**$\beta_1$): We initially hypothesize that there may be a complex relationship between sleep duration on weekdays and sleep troubles.
Long sleep durations could indicate an attempt to compensate for poor quality sleep, potentially related to sleep disturbances.
As such, the prior for this coefficient is centered around zero, allowing the data to indicate the direction of association.

**Weekend Sleep Duration (**$\beta_2$): For sleep duration on weekends, we consider the possibility of a recuperative effect, where catching up on sleep may reduce the likelihood of reporting troubles.
This hypothesis is tentative, hence a prior centered around zero with a conservative standard deviation, permitting data-driven insights into this relationship.

**Snoring Frequency (**$\beta_3$): Frequent snoring is a recognized symptom of sleep disorders like obstructive sleep apnea, which can lead to disrupted sleep.
We posit a positive association between snoring frequency and reported sleep troubles, reflected in the priors for this coefficient.

**Breathing Pause Frequency (**$\beta_4$): Breathing pauses during sleep are indicative of sleep apnea, which is directly related to sleep troubles.
We hypothesize a positive association for this coefficient, which is incorporated into the prior setting.

**Daytime Sleepiness (**$\beta_5$): Excessive daytime sleepiness is often the result of inadequate or disrupted nighttime sleep.
We hypothesize a positive relationship between daytime sleepiness and the reporting of sleep troubles, expecting that those experiencing higher levels of sleepiness will be more likely to report troubles.

@@

# Print the summary table
coefficient_table
```

@tbl-logistic-coefficients presents the key coefficients for understanding the influence of various sleep-related variables on the likelihood of reporting sleep troubles:

**Intercept (**$\beta_0$): The intercept of the model, $\beta_0$, is estimated at -1.396.
This value represents the log odds of reporting sleep troubles when all the predictors are held at zero.
It suggests that, in the absence of any sleep issues or without considering any sleep behaviors, the baseline log odds of reporting sleep troubles to a doctor are negative, indicating a lower probability of reporting sleep troubles.

**Weekday Sleep Duration (**$\beta_1$): For the weekday sleep duration, the coefficient $\beta_1$ is 0.051, suggesting a slight increase in the log odds of reporting sleep troubles for each additional hour of sleep during the weekdays.
This could be indicative of a situation where those who sleep more on weekdays might be doing so due to sleep issues that have led them to consult with a doctor.

**Weekend Sleep Duration (**$\beta_2$): Conversely, the coefficient for weekend sleep duration, $\beta_2$, is -0.131.
This implies that for each additional hour of sleep during the weekends, there is a decrease in the log odds of reporting sleep troubles.
It could be interpreted as those who manage to sleep more on weekends are less likely to report sleep problems, possibly because catching up on sleep may alleviate some of their weekday sleep deficits.

**Snoring Frequency (**$\beta_3$): The coefficient for snoring frequency, $\beta_3$, has a value of 0.032.
While this is a positive value, suggesting that an increase in snoring frequency is associated with an increase in the likelihood of reporting sleep troubles, the effect size is relatively small.

**Breathing Pause Frequency (**$\beta_4$): The coefficient for breathing pause frequency, $\beta_4$, is 0.412, indicating a more substantial positive association with reporting sleep troubles.
This aligns with clinical understanding, as breathing pauses are often associated with sleep disorders like sleep apnea, which can be a significant concern prompting medical consultation.

**Overly Sleep Frequency (**$\beta_5$): Lastly, the coefficient for overly sleep frequency, $\beta_5$, is 0.425, suggesting a strong relationship with the reporting of sleep troubles.
This result is intuitive, as feeling excessively sleepy can be a direct symptom of poor sleep quality or a sleep disorder, leading to discussions with a healthcare provider.

@@

```{=tex}
\begin{align}
\text{logit}(p_i) &= -1.396 + 0.051 \times x_{i1} - 0.131 \times x_{i2} + 0.032 \times x_{i3} + 0.412 \times x_{i4} + 0.425 \times x_{i5} + \varepsilon_i
\end{align}
```
## Predict Sleep Disorder Through Self Reported Patterns

@@

## Model Performance

```{r}
#| echo: false
# Predict probabilities on the training data
predicted_probabilities <- predict(bayesian_logistic_model, type = "response")

# Set a new threshold for classification
threshold <- 0.5  # Adjust this value as needed

# Calculate the predicted classes based on the threshold
predicted_classes <- ifelse(predicted_probabilities > threshold, 1, 0)

# Actual classes from the data
actual_classes <- cleaned_data$ReportedSleepTrouble

# Create a confusion matrix
conf_matrix <- table(Predicted = predicted_classes, Actual = actual_classes)

# Extract components and round them to integer
TP <- as.integer(conf_matrix["1", "1"])  # True Positives
TN <- as.integer(conf_matrix["0", "0"])  # True Negatives
FP <- as.integer(conf_matrix["1", "0"])  # False Positives
FN <- as.integer(conf_matrix["0", "1"])  # False Negatives

# Calculate accuracy and round rates to three decimal places
accuracy <- round(sum(diag(conf_matrix)) / sum(conf_matrix), 3)
TP_rate <- round(TP / (TP + FN), 3)
TN_rate <- round(TN / (TN + FP), 3)
FP_rate <- round(FP / (FP + TN), 3)
FN_rate <- round(FN / (FN + TP), 3)

# Define metrics data frames
counts_df <- data.frame(
  Metric = c("True Positives", "True Negatives", "False Positives", "False Negatives"),
  Count = c(TP, TN, FP, FN)
)

rates_df <- data.frame(
  Metric = c("Accuracy", "True Positive Rate", "True Negative Rate", "False Positive Rate", "False Negative Rate"),
  Rate = c(accuracy, TP_rate, TN_rate, FP_rate, FN_rate)
)

# Create tables with kable and apply styling
counts_table <- kable(counts_df, col.names = c("Metric", "Count"), align = c('l', 'r')) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

rates_table <- kable(rates_df, col.names = c("Metric", "Rate"), align = c('l', 'r')) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r}
#| label: tbl-performance-count
#| tbl-cap: "Count Metrics of the Model Performance"
#| echo: false
# Generate the table for counts
counts_table <- kable(counts_df, 
                       col.names = c("Metric", "Count"), 
                       align = c('l', 'r')) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)

#Print the table
counts_table
```
@tbl-performance-count showcases how well the model predicts reported sleep troubles with counts of true positives and true negatives indicating correct predictions for the presence and absence of sleep trouble, respectively. Conversely, false positives and false negatives highlight where the model predicts incorrectly, either by overestimating (predicting sleep trouble where there is none) or underestimating (failing to detect reported sleep trouble) the condition. These counts are essential for assessing the model's performance accuracy, balancing the sensitivity to detect actual cases against the specificity to dismiss non-cases.

```{r}
#| label: tbl-performance-rate
#| tbl-cap: "Rate Metrics of the Model Performance"
#| echo: false
# Generate the table for rates
rates_table <- kable(rates_df, 
                     col.names = c("Metric", "Rate"), 
                     align = c('l', 'r')) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)

#Print the table
rates_table
```

@tbl-performance-rate shows the performance indicators of the model. The accuracy stands at 75.3%, denoting the proportion of predictions that the model got right. The true positive rate is 15.3%, reflecting the model’s efficiency in correctly predicting actual instances of sleep trouble. Impressively, the true negative rate is 96.4%, indicating the model's strong capability to recognize individuals without sleep trouble. Conversely, the false positive rate is minimal at 3.6%, suggesting that the model seldom incorrectly flags sleep trouble where there is none. However, the model’s potential area for improvement is highlighted by the false negative rate of 84.7%, which points to a substantial number of actual sleep trouble cases that the model fails to detect. These metrics are instrumental for evaluating the model's precision and guiding further refinements.

@@

## Advantages of Self-Reported Data in Sleep Health

Using self-reported patterns to predict sleep troubles offers a valuable, accessible means for identifying individuals who may require medical intervention. This approach leverages personal sleep logs and questionnaires, which are cost-effective and can be implemented on a large scale, enhancing the reach of health monitoring. Self-reported data provide insights into an individual's sleep duration, frequency of disturbances like snoring and breathing pauses, and subjective feelings of sleepiness, all of which are crucial for diagnosing sleep disorders.

## Impact of Sleep Patterns on Reported Sleep Troubles {#sec-sleep-pattern-impact}

The analysis presented in this paper underscores the complex interplay between various sleep behaviors and the likelihood of individuals reporting sleep troubles. Our model's findings align with existing literature suggesting that both insufficient sleep duration and poor sleep quality are significant predictors of sleep disorders [@State_2024]. In particular, the rates of false negatives indicate a substantial portion of individuals experiencing sleep issues that the model failed to capture, possibly reflecting the nuanced and subjective nature of sleep disturbances.

## The Role of Breathing Irregularities and Daytime Sleepiness {#sec-breathing-daytime-sleep}

Breathing irregularities and excessive daytime sleepiness emerged as strong predictors in our model, pointing towards conditions such as sleep apnea, which remains underdiagnosed in the general population [@Foundation_2021]. The high false negative rate may partly result from such conditions' subtler signs, which individuals may not report or may attribute to other causes.

## Assessing Model Performance and Clinical Implications {#sec-model-performance}

While the model demonstrated robust true negative rates, indicating a strong ability to correctly identify individuals without sleep troubles, the high false negative rate poses significant concerns. Clinically, this could lead to missed diagnoses and a delay in treatment for those with unreported sleep disorders. Future iterations of the model might benefit from incorporating additional predictors or employing different statistical techniques to enhance sensitivity.

## Limitations and Future Directions {#sec-limitations-future-directions}

### Limitations {#sec-limitations}

A significant limitation of this analysis was the sole reliance on self-reported data, which can introduce response bias and affect the accuracy of the findings. Additionally, the dataset may not adequately represent the entire population, as some demographics may be underrepresented. An inherent challenge also lay in the absence of a validation set, which could lead to an overfit model that doesn't generalize well to new, unseen data.

### Future Directions {#sec-future-directions}

For future research, it would be beneficial to conduct longitudinal studies to understand the temporal aspects of sleep patterns better and their long-term health impacts. Utilizing a separate validation dataset would help in assessing the model's predictive power and in refining the classification threshold to balance the trade-off between sensitivity and specificity effectively. Incorporating objective data sources, such as wearable technology that provides sleep quality indicators, could greatly enhance the robustness of the findings. Additionally, employing advanced machine learning algorithms might reveal complex non-linear interactions and patterns within the predictors, offering richer insights than those obtainable from traditional logistic regression analysis. Emphasizing the development of predictive models that are both interpretable and accurate will be key to advancing our understanding of sleep health.

\newpage
# Appendix {.unnumbered}

@@

From @fig-ppcheckandposteriorvsprior, from the left plot the alignment of the observed data (solid black line) with the simulations from the model (lighter blue lines) suggests that the model captures the essential trends and variability in the data. This congruence indicates a model that adequately reflects the observed phenomena and provides assurance that the predictive capacity of the model is robust for the data in question.

The right plot displays the comparison between the posterior and prior distributions which offers insights into the influence of the data on our initial assumptions. The posterior distributions remain close to the priors, suggesting that the priors were well-calibrated and that the data has refined, rather than overturned, these initial beliefs. The spread of the posterior distributions signifies the inherent uncertainty and variability within the model, which is a fundamental aspect of statistical modeling and Bayesian analysis.

\newpage
